---
title: "Assignment 6"
author: "Darshan Patel"
date: '2022-07-24'
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
# Add any packages you want in this chunk:
library("olsrr")
```

We are going to look at the property prices for Orange County. From there we are going to try and predict the sales price by linear regression.

# Importing Data 

Bring in the data and make sure the data types are correct. If not, make the proper changes. The file is located within this project. *data/prop_prices_reduced.csv*

```{r}
data <- read.csv("/Users/darshanpatel/Documents/GitHub/ECO5445SU22/Assignment06/data/prop_prices_reduced.csv", header = TRUE, stringsAsFactors = FALSE)
```


# Plotting

Plot histograms for all variables. Additionally, add scatterplots for the relationships between all quantitative variables.

```{r}
hist(data$sale_def)
hist(data$bed)
hist(data$bath)
hist(data$area_heated)
hist(data$area)
hist(data$dist_cbd)
hist(data$dist_lakes)
hist(data$pool)

plot(data$area_heated, data$sale_def)
plot(data$area, data$sale_def)
plot(data$dist_cbd, data$sale_def)
plot(data$dist_lakes, data$sale_def)
plot(data$area_heated, data$area)
plot(data$dist_cbd, data$area)
plot(data$dist_lakes, data$area)

```

# Summary Statistics

Provide basic summary statistics for univariate analysis. Also, provide the correlation between all the quantitative variables.

```{r}
summary(data)

cor(data$area_heated, data$sale_def)
cor(data$area, data$sale_def)
cor(data$dist_cbd, data$sale_def)
cor(data$dist_lakes, data$sale_def)
cor(data$area_heated, data$area)
cor(data$dist_cbd, data$area)
cor(data$dist_lakes, data$area)

```

# Regression Analysis

Run a regression with all the variables included. Print results of the regression.

```{r}
Model <- lm(data$sale_def ~ data$bed + data$bath + data$area_heated + data$area + data$dist_cbd + data$dist_lakes + data$pool)

print(Model)

summary(Model)

```

Which of the variables tested significant at the 95% level? Looking at the results and answering outside of the chunk is sufficient.

## Based on the summary statistics of the linear regression, the variables that tested significant at the 95% confidence level are the pool and bed independent variables since their p value is .05, which equates to a 95% confidence level.

## Evaluating the model

As is, are any of the Gauss-Markov assumptions violated? If so, which ones? How can you fix the issues?

```{r}
ols_vif_tol(Model)
# based on the VIF factor, multicollinearity does not seem to be an issue in regards to the assumptions.
#Assuming no other assumptions are violated since random disturbance is unknown and other assumptions rely on random disturbance

```

## New Model

Based off of your findings in the previous section, make changes to the variables, the functional form, etc.

```{r}
# No changes to be made.
```


# Prediction

Based on the following inputs, predict the deflated sales price:

- 2 bed
- 2 bath
- area_heated = 1223
- area = 9750
- dist_cbd = 19368
- dist_lakes = 490
- no pool

```{r}
Price = (-8.492e+04*2) + (5.449e+04*2) + (2.439e+02*1223) + (3.548e+00*9750) + (-5.848e+00*19368) + (9.651e-01*490) + (-2.776e+04*0)

print(Price)
```
